---
output: pdf_document
---

------------------------------------------------------------------------

## Installation

```{r}
# The following line causes a compliation issue when knitting, uncomment if you run yourself
# install.packages('tidyverse')
# install.packages('dplyr')
library(tidyverse)
library(haven)
library(dplyr)
```

## Pre-process dataset

```{r}
# Treatment Variable: H4WP3 (if the biological mother has ever spent time in jail or prison)
# Outcome Variable: H4ED1 (respondent’s high school graduation status)

# CHANGE THIS TO WHERE YOUR WAVE 4 DATASET IS. THE WAVE 4 FOLDER SHOULD BE NAMED 'Wave4'
wave4_dir <- "Wave4/"

concat <- function(s1, s2) {
  return (paste(s1, s2, sep=""))
}

# w4inhome_dvn.sav is the main datatable that contains our treatment and outcome variable
inhome_df <- read_sav(concat(wave4_dir, "w4inhome_dvn.sav")) 
birth_df <- read_sav(concat(wave4_dir, "w4birth_dvn.sav"))
segment_df <- read_sav(concat(wave4_dir, "w4segment_dvn.sav"))

# join against these table because they could have useful information
# keep rows with unique AIDs only
birth_df <- birth_df %>% distinct(AID, .keep_all = TRUE)
segment_df <- segment_df %>% distinct(AID, .keep_all = TRUE)

combined_df <- inhome_df %>% 
  left_join(birth_df, by = "AID") %>%
  left_join(segment_df, by = "AID")


relevant_df = combined_df %>% 
  filter(H4WP3 <= 1) %>% # Filter out refused or don't know for H4WP3
  filter(H4ED1 <= 4) %>% # Filter our don't know for H4ED1
  mutate(H4WP4 = case_when(H4WP4 >= 97 ~ NA, .default = H4WP4),
         H4WP5 = case_when(H4WP5 >= 94 ~ NA, .default = H4WP5),
         H4WP24 = case_when(H4WP24 == 7 ~ NA, .default = H4WP24),
         H4EC1 = case_when(H4EC1 >= 96 ~ NA, .default = H4EC1),
         H4ED7 = case_when(H4ED7 >= 5 ~ NA, .default = H4ED7),
         H4EC7 = case_when(H4EC7 >= 96 ~ NA, .default = H4EC7),
         H4EO7 = case_when(H4EO7 > 4 ~ NA, .default = H4EO7)) %>% #Change Don't Know/Missing/Refuse values to NA rather than numerical code
  select(H4WP3, H4WP4, H4WP5, H4WP24, H4OD4, H4EC1, H4EC7, H4LM28, H4ED7, H4EO7, BIO_SEX4, H4ED1) 

relevant_df <- relevant_df %>% 
  rename(Mother.Incarcerated = H4WP3, 
         Mother.Incarcerated.Times = H4WP4,
         Mother.Incarcerated.Respondent.Age = H4WP5,
         Mother.Relationship.Respondent = H4WP24,
         US.Citizen = H4OD4,
         Household.Income = H4EC1,
         Household.Assets = H4EC7,
         Family.Support.Education = H4ED7,
         Family.Responsibilities = H4LM28,
         Neighborhood.Safety = H4EO7,
         Sex = BIO_SEX4,
         Highschool.Graduated = H4ED1)

# Print colnames, look up colnames on 
# https://addhealth.cpc.unc.edu/documentation/codebook-explorer/#/
# To see if they are relevant
colnames(combined_df)

# Print preview for dataset
relevant_df

```

## Overview of Data Pre-processing

### Treatment and Outcome Variables

Our treatment variable is H4WP3 (if the biological mother has ever spent time in jail or prison). Our outcome variable is H4ED1 (respondent's high school graduation status).

$A_i = \begin{cases}
            0 & \text{ if biological mother never spent time in prison}\\
        1 & \text{ if biological mother ever spent time in prision }
        \end{cases}$

$Y_i = \begin{cases}
            0 & \text{ if did not received a high school diploma or equivalent}\\
        1 & \text{ if received a high school diploma or equivalent }
        \end{cases}$

### Tables

We use the Wave 4 dataset from the ADD Health longitudinal study. Our treatment and outcome variables are found in the 'in-home' dataset within W4. Within this dataset, each respondent is classified by an AID. Since there are data present in order tables that we may need to adjust for, such as sex, partner relationships, birth information, etc., we join the 'in-home' table with the 'birth_dvn' and 'w4segment_dvn' tables also found in Wave 4 on AID to incorporate additional relevant information into our analysis.

### Data Adjustment

For our treatment variable, we drop rows where our treatment variable is not well-defined (respondent entered don't know or did not respond).

For our outcome variable, we grouped 'finished high school with a diploma' and 'earned a high school equivalency degree' as one and 'earned a certificate of attendance or certificate of completion' and 'did not receive a high school diploma, equivalency degree (GED), or other certificate' as 0. Furthermore, we drop rows whose graduation status is 'don't know.'

For confounding and other relevant variables, we changed the encoded numerical value for 'don't know', 'refused', or other similar missing data to NA values.

### Description of Columns

Visit <https://addhealth.cpc.unc.edu/documentation/codebook-explorer/> for a more detailed description of the columns.

**Mother.Incarcerated** (H4WP3) [Treatment Variable]: (Has/did) your biological mother ever (spent/spend) time in jail or prison?

**Mother.Incarcerated.Times** (H4WP4): How many times (has/did) your biological mother (spent/spend) time in jail or prison?

**Mother.Incarcerated.Respondent.Age** (H4WP5): How old were you when your biological mother went to jail or prison (the first time)?

**Mother.Relationship.Respondent** (H4WP24): How close do you feel to your [mother figure]?

**US.Citizen** (H4OD4): Were you born a U.S. citizen?

**Household.Income** (H4EC1): Thinking about your income and the income of everyone who lives in your household and contributes to the household budget, what was the total household income before taxes and deductions in {2006/2007/2008}? Include all sources of income, including non-legal sources.

**Household.Assets** (H4EC7): What is your best estimate of the total value of your assets and the assets of everyone who lives in your household and contributes to the household budget? Include all assets, such as bank accounts, retirement plans and stocks. Do not include equity in your home.

**Family.Support.Education** (H4ED7): In the past 12 months, have any relatives, including your parents or in-laws, helped you out by paying some of your educational expenses, such as tuition or books?

**Family.Responsibilities** (H4LM28): Indicate how much you would agree or disagree with this statement: Family responsibilities have interfered with my ability to work.

**Neighborhood.Safety** (H4EO7): How safe did you feel when you were in the sample member's/respondent's neighborhood?

**Sex** (BIO_SEX4): Respondent's Gender

**Highschool.Graduated** (H4ED1) [Outcome Variable]: What is your high school graduation status?

### Consistency and Positivity

**Consistency**

This assumption seems reasonable for our analysis since the outcome (high school graduation status) and treatment (whether the biological mother was incarcerated) are clearly defined
However, discrepancies in misreporting or recall bias could affect results as respondents may inaccurately recall their mother’s incarceration or graduation status. 
Overall, assuming accurate reporting, the consistency assumption holds and is unlikely to pose significant problems in our analysis.

**Positivity**

To check positivity, we need to ensure overlap between the treated and control groups. This involves checking whether the propensity scores for the treated group overlap with those of the control group.

```{r}
library(MatchIt)

ps_model <- glm(Mother.Incarcerated ~ Household.Income + Neighborhood.Safety + Family.Support.Education,data = relevant_df, family = "binomial")

relevant_df <- relevant_df %>% drop_na(Household.Income, Neighborhood.Safety, Family.Support.Education)

relevant_df$propensity_score <- predict(ps_model, type = "response")

```

Now, we Visualize Overlap of Propensity Scores across the two groups through density plot.

```{r}
ggplot(relevant_df, aes(x = propensity_score, fill = as.factor(Mother.Incarcerated))) +
  geom_density(alpha = 0.5) +labs(title = "Propensity Score Distribution by Treatment Group",x = "Propensity Score",fill = "Mother Incarcerated") +theme_minimal()

```

Looking at the above plot, Substantial overlap was observed between the two groups, satisfying the positivity condition.

To ensure no positivity violations, we restricted the analysis to individuals whose propensity scores fall within the overlapping range.

```{r}
min_score <- max(
  min(relevant_df$propensity_score[relevant_df$Mother.Incarcerated == 1]),
  min(relevant_df$propensity_score[relevant_df$Mother.Incarcerated == 0])
)
max_score <- min(
  max(relevant_df$propensity_score[relevant_df$Mother.Incarcerated == 1]),
  max(relevant_df$propensity_score[relevant_df$Mother.Incarcerated == 0])
)

relevant_df <- relevant_df %>%filter(propensity_score >= min_score & propensity_score <= max_score)

plot2 <- ggplot(relevant_df, aes(x = propensity_score, color =factor(Mother.Incarcerated))) +geom_density() +ggtitle("Second Plot: Propensity Score Density After Filtering") +
  labs(color = "Mother Incarcerated") +theme_minimal()

print(plot2)

```

A second density plot confirms that the filtered dataset maintains substantial overlap in propensity scores.The positivity assumption is meet after filtering the data to the region of common support. Substantial overlap in propensity scores confirms that all groups have a chance of receiving either treatment condition. 